---
title: "Statistics and Quantitative Methods (S2)"
subtitle: "Week 5"
author: "Dr Stefano Coretta"
institute: "University of Edinburgh"
date: "2023/02/14"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css:
      - ../xaringan-themer.css
      - ../custom.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "../macros.js"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=7, fig.height=5, fig.retina=3,
  out.width = "60%", fig.align = "center",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
knitr::opts_knit$set(root.dir = here::here())

library(xaringanExtra)
use_xaringan_extra(c("panelset", "tachyons", "freezeframe"))

library(tidyverse)
theme_set(theme_light())
library(brms)
library(extraDistr)
library(ggdist)
library(glue)

options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(8, "Dark2"))
options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(8, "Dark2"))
options(show.signif.stars = FALSE)
my_seed <- 8878
```

```{r read-data}
polite <- read_csv("data/polite.csv") %>%
  mutate(
    attitude = recode_factor(attitude, "inf" = "informal", "pol" = "polite")
  )

alb_vot <- read_csv("data/alb_vot.csv") %>%
  mutate(
    vot = (voi_onset - release) * 1000
  )

alb_vot_vl <- alb_vot %>%
  filter(
    label %in% c("k", "p", "t"),
    vot < 20
  )
```

## Summary

---

layout: true

## Comparing groups

---

```{r hnr}
polite %>%
  ggplot(aes(attitude, HNRmn, fill = attitude)) +
  geom_violin() +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(
    y = "Harmonics-to-noise ratio (dB)",
    title = "Harmonics-to-noise ratio is Korean speakers"
  )
```

---

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
- We want to estimate the probability distribution of harmonics-to-noise ratio (HNR) in Korean speakers.

  - Let's assume it is a Gaussian probability distribution.
  
  - So, `\(\text{HNR} \sim Gaussian(\mu, \sigma)\)`.
]

--

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
- But we also want to estimate the difference in the mean ($\mu$) in the informal vs polite attitude conditions!
  
  - In other words, we need to allow the model to estimate $\mu$ based on attitude (informal vs polite).
]

---

.f3[
$$\text{HNR} \sim Gaussian(\mu, \sigma)$$
]

--

.f3[
$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol}$$
]

--

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
- $\beta_0$: mean HNR when [attitude = informal].

- $\beta_1$: difference of mean HNR between [attitude = polite] and [attitude = informal].

  - $\beta_1 = \mu_{pol} - \mu_{inf}$
  
  - **`\(\beta_1\)` is the effect of [attitude = polite] on mean HNR relative to the baseline mean HNR when [attitude = informal]**.

- $attitude_{pol}$: is 0 for [attitude = informal] and 1 for [attitude = polite].
]

---

.f3[
$$\text{HNR} \sim Gaussian(\mu, \sigma)$$
]

.f3[
$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol}$$
]

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
**For [attitude = informal]**, $\mu$ is:

$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol} = \beta_0 + \beta_1 \cdot 0 = \beta_0$$

**For [attitude = polite]**, $\mu$ is:

$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol} = \beta_0 + \beta_1 \cdot 1 = \beta_0 + \beta_1$$
]

---

.f3[
$$\text{HNR} \sim Gaussian(\mu, \sigma)$$
]

.f3[
$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol}$$
]

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
**For [attitude = informal]**, $\mu$ is:

$$\mu_{\text{(attitude = informal)}} = \beta_0$$

**For [attitude = polite]**, $\mu$ is:

$$\mu_{\text{(attitude = polite)}} = \beta_0 + \beta_1$$
]

---

.f3[
$$\text{HNR} \sim Gaussian(\mu, \sigma)$$
]

.f3[
$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol}$$
]

--

<br>

.f3[
$$\beta_0 \sim Gaussian(\mu_0, \sigma_0)$$
]

.f3[
$$\beta_1 \sim Gaussian(\mu_1, \sigma_1)$$
]

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
- We estimate the probability distribution of mean HNR when [attitude = informal].

- We estimate the probability distribution of the difference in mean HNR between [attitude = polite] and [attitude = informal]. 
]

---

```{r hnr-bm, echo=FALSE}
hnr_bm <- brm(
  HNRmn ~ attitude,
  data = polite,
  backend = "cmdstanr",
  cores = 4,
  file = "./data/cache/hnr_bm",
  seed = my_seed
)
```

```{r hnr-bm-code, eval=FALSE, echo=TRUE}
# Attach the brms package
library(brms)

# Run a Bayesian model
vot_bm <- brm(
  # This is the formula of the model.
  HNRmn ~ 1 + attitude,
  # This is the probability distribution family.
  family = gaussian(),
  # And the data.
  data = polite
)
```

---

.f2.center[
`HNRmn ~ 1 + attitude`
]

**Read as**: Model HNR values (`HNRmn`) as a function of (`~`) the mean (`1`) and attitude (`attitude`).

--

.f2.center[
`HNRmn ~ attitude`
]

We can omit the `1` (it is implied by default when there are other terms in the formula)! **You can read as**: Model HNR values (`HNRmn`) as a function of (`~`) attitude (`attitude`).

--

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
- The variable to the left of `~` is called the **outcome variable** (aka response variable, dependent variable).

- The variable(s) to the right of `~` are called **predictor variables** (aka independent variables).
]

---

.f2.center[
`HNRmn ~ attitude`
]

.bg-washed-green.b--dark-green.ba.bw2.br3.shadow-5.ph4.mt2[
😱 But... `attitude` is not a numeric predictor! 
]

```{r att-col}
polite
```


---

.f3[
$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol}$$
]

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
The formula works with numeric variables.

- But `attitude` is a categorical variable.

- 🤔 Think about it: what does it mean to multiply $\beta_1$ by "informal" or "polite"?
]

--

.bg-washed-green.b--dark-green.ba.bw2.br3.shadow-5.ph4.mt2[
Let's talk about coding of categorical variables!
]

<br>

**NOTE**: What follows is for you how to understand how coding works, but remember that this is done automatically by R for you so you never have to do it by hand!!!

---

layout: false
layout: true

## Coding categorical predictors

---

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
**Categorical predictors can be coded using numbers.**

There are two common types of coding systems:

- **Treatment** coding (this week's focus).

- **Sum** coding.
]

--

.bg-washed-green.b--dark-green.ba.bw2.br3.shadow-5.ph4.mt2[
As with anything else in stats, **naming of coding systems is not an established matter** and the same coding can have different names, and vice versa the same name could refer to different systems.

For an excellent overview, see <https://debruine.github.io/faux/articles/contrasts.html>.
]

---

**Treatment** coding uses `0` and `1` to code categorical predictors.

<br>

.f3[
|               | attitude_pol |
| ------------- | -------:     |
| informal      | 0            |
| polite        | 1            |
]

---

```{r att-coding}
polite %>%
  mutate(
    attitude_pol = ifelse(attitude == "informal", 0, 1)
  ) %>%
  select(HNRmn, attitude, attitude_pol)
```

**NOTE**: Remember that this is done automatically by R under the hood for you so you never have to do it by hand!!!

---

.f3[
|               | attitude_pol |
| ------------- | -------:     |
| informal      | 0            |
| polite        | 1            |
]

<br>

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
$$\mu_{\text{(attitude = informal)}} = \beta_0 + \beta_1 \cdot attitude\_pol = \beta_0 + \beta_1 \cdot 0 = \beta_0$$

$$\mu_{\text{(attitude = polite)}} = \beta_0 + \beta_1 \cdot attitude\_pol = \beta_0 + \beta_1 \cdot 1 = \beta_0 + \beta_1$$
]

---

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
With **treatment coding**, the first level in the predictor is the **reference level** (in `attitude`, it is `informal`).

The other level is compared against the reference level (`polite` vs `informal`).
]

--

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
**Level order**

- By default, the ordering of levels in a categorical predictor is based on alphabetical order.

- But you can specify the order manually using the `factor()` function. (You will see how in the tutorial).
]

---

.f3[
$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol}$$
]

<br>

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
To reiterate from above:

- $\beta_0$ corresponds to the mean NHR when [attitude = informal], because `informal` is the reference level of `attitude`.

- $\beta_1$ corresponds to the **DIFFERENCE** between $\beta_0$ and mean HNR when [attitude = polite], because `polite` is the other level in `attitude`.

  - In other words, we are comparing mean HNR when [attitude = polite] vs [attitude = informal].
]

--

.bg-washed-green.b--dark-green.ba.bw2.br3.shadow-5.ph4.mt2[
For now, things are easy because we only have two levels (informal vs polite), but we will see an example with three levels later.
]

---

layout: false
layout: true

## Modelling HNR by attitude

---

```{r hnr-bm-code-2, eval=FALSE, echo=TRUE}
vot_bm <- brm(
  # This is the formula of the model. We can omit `1`.
  HNRmn ~ attitude,
  # This is the probability distribution family.
  family = gaussian(),
  # And the data.
  data = polite
)
```

--

$$\text{HNR} \sim Gaussian(\mu, \sigma)$$

$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol}$$

$$\beta_0 \sim Gaussian(\mu_0, \sigma_0)$$

$$\beta_1 \sim Gaussian(\mu_1, \sigma_1)$$

$$\sigma \sim TruncGaussian(\mu_2, \sigma_2)$$

--

.bg-washed-green.b--dark-green.ba.bw2.br3.shadow-5.ph4.mt2[
We need to estimate $\mu_0$, $\sigma_0$, $\mu_1$, $\sigma_1$, $\mu_2$, $\sigma_2$.
]

---

```{r hnr-bm-summ}
summary(hnr_bm)
```

---

```{r hnr-bm-summ-2}
hnr_bm_summ <- capture.output(summary(hnr_bm))
cat(hnr_bm_summ[8:11], sep = "\n")
```

--

<br>
<br>

.pull-left[
- **Intercept**: $\mu_{\text{attitude = informal}} = \beta_0$.

- **Estimate**: $\mu_0 = 16.27$ dB.

- **Est.Error**: $\sigma_0 = 0.16$ dB.
]

.pull-right[
$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol}$$

.purple[
$$\beta_0 \sim Gaussian(16.27, 0.16)$$
]

$$\beta_1 \sim Gaussian(\mu_1, \sigma_1)$$
]

--

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
When [attitude = informal], the probability distribution of $\mu$ is the probability distribution of $\beta_0$, which is $Gaussian(16.27, 0.16)$.
]

---

```{r hnr-int-p}
hnr_bm_draws <- as_draws_df(hnr_bm)
int_dens_l <- density(hnr_bm_draws$b_Intercept, adjust = 1.5)
int_dens <- tibble(
  x = int_dens_l$x,
  y = int_dens_l$y
)

hnr_int_p <- int_dens %>%
  ggplot(aes(x, y)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(n.breaks = 6) +
  labs(
    title = expression(Probability~distribution~of~beta[0]),
    x = expression(beta[0]), y = "Probability density"
  ) +
  ylim(0, 3)
hnr_int_p
```

---

```{r hnr-int-p-2}
ci_low <- 15.97
ci_hi <- 16.59
label_y <- 2.75

hnr_int_p +
  geom_ribbon(
    aes(x = ifelse(x >= ci_low & x <= ci_hi, x, NA), ymin = 0, ymax = y),
    fill = "#9970ab",
    alpha = 0.4
  ) +
  annotate(
    "segment",
    x = ci_low, xend = ci_hi, y = label_y, yend = label_y,
    arrow = arrow(ends = "both", angle = 90, length = unit(.2, "cm")),
    size = 1
  ) +
  annotate(
    "segment",
    x = ci_low, xend = ci_low, y = 0, yend = label_y,
    linetype = "dashed"
  ) +
  annotate(
    "segment",
    x = ci_hi, xend = ci_hi, y = 0, yend = label_y,
    linetype = "dashed"
  ) +
  annotate("label", x = ci_low + (ci_hi-ci_low)/2, y = label_y, label = "95% CrI")
```

--

There is a 95% probability that mean HNR when [attitude = informal] is between `r ci_low` and `r ci_hi` dB.


---

```{r nhr-bm-summ-3}
hnr_bm_summ <- capture.output(summary(hnr_bm))
cat(hnr_bm_summ[8:11], sep = "\n")
```

<br>
<br>

.pull-left[
- **attitudepolite**: $\beta_1$.

- **Estimate**: $\mu_0 = 1.25$ dB.

- **Est.Error**: $\sigma_0 = 0.23$ dB.
]

.pull-right[
$$\mu = \beta_0 + \beta_1 \cdot attitude_{pol}$$

$$\beta_0 \sim Gaussian(16.27, 0.16)$$

.purple[
$$\beta_1 \sim Gaussian(1.25, 0.23)$$
]
]

--

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
When [attitude = polite], the probability distribution of the difference between $\mu_{\text{attitude = polite}}$ and $\mu_{\text{attitude = informal}}$ is the probability distribution of $\beta_1$, which is $Gaussian(1.25, 0.23)$.
]

---

```{r hnr-att-p}
att_dens_l <- density(hnr_bm_draws$b_attitudepolite, adjust = 1.5)
att_dens <- tibble(
  x = att_dens_l$x,
  y = att_dens_l$y
)

hnr_att_p <- att_dens %>%
  ggplot(aes(x, y)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(n.breaks = 6) +
  labs(
    title = expression(Probability~distribution~of~beta[1]),
    x = expression(beta[1]), y = "Probability density"
  ) +
  ylim(0, 2)
hnr_att_p
```

---

```{r hnr-att-p-2}
ci_low <- 0.8
ci_hi <- 1.7
label_y <- 2

hnr_att_p +
  geom_ribbon(
    aes(x = ifelse(x >= ci_low & x <= ci_hi, x, NA), ymin = 0, ymax = y),
    fill = "#9970ab",
    alpha = 0.4
  ) +
  annotate(
    "segment",
    x = ci_low, xend = ci_hi, y = label_y, yend = label_y,
    arrow = arrow(ends = "both", angle = 90, length = unit(.2, "cm")),
    size = 1
  ) +
  annotate(
    "segment",
    x = ci_low, xend = ci_low, y = 0, yend = label_y,
    linetype = "dashed"
  ) +
  annotate(
    "segment",
    x = ci_hi, xend = ci_hi, y = 0, yend = label_y,
    linetype = "dashed"
  ) +
  annotate("label", x = ci_low + (ci_hi-ci_low)/2, y = label_y, label = "95% CrI")
```

--

There is 95% probability that the difference in mean HNR between $\mu_{\text{attitude = polite}}$ and $\mu_{\text{attitude = informal}}$ is between `r ci_low` and `r ci_hi` dB.

